Icicle: a single-pass query language

### 2016-02-24 2016icicle/section/01-Introduction.tex
break everything

For feature generation for machine learning, our queries have some quite specific restrictions. \ben{The following describes properties that that queries and data have intrinsically, not \emph{restrictions} placed on them}.
Queries only need to read from one `table', for example the stocks table with company codes, dates and open and close prices for each day.
Each query is implicitly partitioned by the entity, the company code in this case, and performs an aggregate over that company's entries, sorted by date.
In this way, our query language is actually a streaming dataflow language with only one input stream and clock.
By making use of these restrictions, we are able to generate more efficient code than a general query language.

>>>>

\amos{Yes, I had some concrete examples earlier but they turned out to be invalid. How about something along these lines?}
\begin{code}
SELECT open FROM stocks;

SELECT open / (SELECT AVG(open) FROM stocks)
FROM stocks;

SELECT open / (SELECT AVG(open) FROM stocks),
       open - (SELECT MIN(open) FROM stocks)
FROM stocks;

SELECT open / avgopen, open - minopen
FROM stocks,
(SELECT AVG(open) as avgopen,
        MIN(open) minopen FROM stocks) agg;

SELECT (SELECT AVG(open) FROM stocks)
       (SELECT MIN(open) FROM stocks)
FROM ();

SELECT AVG(open) as avgopen,
       MIN(open) minopen FROM stocks
\end{code}

For feature generation for machine learning, our queries have some quite specific restrictions.
\ben{The following describes properties that that queries and data have intrinsically, not \emph{restrictions} placed on them}.
\amos{I don't understand what this means. Is an SQL query with cross product, join, etc not a query?}

Queries only need to read from one `table', for example the stocks table with company codes, dates and open and close prices for each day.
Each query is implicitly partitioned by the entity, the company code in this case, and performs an aggregate over that company's entries, sorted by date.
In this way, our query language is actually a streaming dataflow language with only one input stream and clock.
By making use of these restrictions, we are able to generate more efficient code than a general query language.

### 2016-02-24 2016icicle/section/02-Source.tex
break everything

First we define a function to compute the sum of some input data.
While traditionally sum would take just an input stream of numbers, we separate this into two distinct parameters.
The first parameter is the clock for the input stream, which defines the parts of the input stream to sample.
This can be thought of as a stream of boolean flags, where true denotes that the input stream should be sampled at this time.
The second parameter is the actual value to sample at the clock.
In Icicle this has type $\Elm{\NN}$, meaning it is available as a number at any part of the stream.
\ben{Maybe hold off the description of @Element@ and @Aggregate@ until later so we can introduce them both at the same time. At this point in the text it's not clear why we need the modal types.}

@@

The result of $\mi{sum}$ above has type $\Agg{\NN}$.

\ben{Does the concrete syntax support this annotation, eg:}
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@function@ 
$\mi{sum}[c]~(v~:~\NN) : \Agg{\NN}$    \\
\> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$. \\
\end{tabbing}

@@

The judgment $\Bigstep{f}{x}{v}$ takes the list of defined functions, $f$, and the expression to evaluate, $x$, and returns the result value $v$.

@@

For example, for the $\mi{opensum}$ query we would have two evaluations:
\begin{tabbing}
MMMM \= M \= \kill
$\mi{clock}_1$ \> $=$ \> $@Elements@~[@T@]$ \\
$\mi{open}_1$  \> $=$ \> $@Elements@~[10]$ \\
$\Bigstep{\mi{sum}}{\mi{opensum}[c~:=~\mi{clock}_1,~\mi{open}~:=~\mi{open}_1]}{@Scalar@~10} $ \\
\\
$\mi{clock}_2$ \> $=$ \> $@Elements@~[@T@,~@T@]$ \\
$\mi{open}_2$  \> $=$ \> $@Elements@~[11,~12]$ \\
$\Bigstep{\mi{sum}}{\mi{opensum}[c~:=~\mi{clock}_2,~\mi{open}~:=~\mi{open}_2]}{@Scalar@~23} $
\end{tabbing}

>>>>

First we define a function to compute the sum of some input data.
While traditionally sum would take just an input stream of numbers, we separate this into two distinct parameters.
The first parameter is the clock for the input stream, which defines the parts of the input stream to sample.
This can be thought of as a stream of boolean flags, where true denotes that the input stream should be sampled at this time.
The second parameter is the actual value to sample at the clock.
In Icicle this has type , meaning it is available as a number at any part of the stream.
\ben{Maybe hold off the description of @Element@ and @Aggregate@ until later so we can introduce them both at the same time. At this point in the text it's not clear why we need the modal types.}

@@

The result of $\mi{sum}$ above has type .

\ben{Does the concrete syntax support this annotation, eg:}
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@function@ 
$\mi{sum}[c]~(v~:~\NN) : $    \\
\> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$. \\
\end{tabbing}

@@

@@

For example, for the $\mi{opensum}$ query we would have two evaluations:
\begin{tabbing}
MMMM \= M \= \kill
$\mi{clock}_1$ \> $=$ \> $@Elements@~[@T@]$ \\
$\mi{open}_1$  \> $=$ \> $@Elements@~[10]$ \\
\\
$\mi{clock}_2$ \> $=$ \> $@Elements@~[@T@,~@T@]$ \\
$\mi{open}_2$  \> $=$ \> $@Elements@~[11,~12]$ \\
\end{tabbing}

### 2016-02-24 2016icicle/section/03-Types.tex
break everything

The judgment $\WrapMode{\tau}{m}{\tau'}$, corresponds to unboxing the a function argument of type $m$ and reboxing the function return type $\tau$ into modality $m$, with $\tau'$ as the boxed return type.
The rule (WrapModeArrow) simply recurses down through the function arguments, applying the judgment to the result.
The rule (WrapElement) applies when reboxing an @Element@ result as an @Element@, and similarly for (WrapAggregate).
For (WrapPure) the result type is a @Pure@, so it can be boxed to any mode.
If these rules do not apply, for example when an @Aggregate@ result is attempted to be reboxed as an @Element@, the function application is an error.

The judgment $\WrapApp{\tau}{\phi\ldots}{\tau'}$ attempts to apply the function type $\tau$ to the argument list $\phi\ldots$, performing unboxing as necessary.
When the argument list is empty and the function type is a nullary function or simple value type, (WrapAppFinished) simply returns the result type.

@@

First, look at an example that would require two passes over the data: maybe @filter open > mean open in count@.
In this case, $\mit{open}$ has type $\Elm{\NN}$ while $\mit{mean~open}$ has type $\Agg{\NN}$.
The primitive $>$ has type $\NN \to \NN \to \NN$ and can be boxed to either @Element@ or @Pure@, but not both.
This means that $\mit{open} > \mit{mean open}$ is a type error.

>>>>

The rule (WrapModeArrow) simply recurses down through the function arguments, applying the judgment to the result.
The rule (WrapElement) applies when reboxing an @Element@ result as an @Element@, and similarly for (WrapAggregate).
For (WrapPure) the result type is a @Pure@, so it can be boxed to any mode.
If these rules do not apply, for example when an @Aggregate@ result is attempted to be reboxed as an @Element@, the function application is an error.

When the argument list is empty and the function type is a nullary function or simple value type, (WrapAppFinished) simply returns the result type.

@@

First, look at an example that would require two passes over the data: maybe @filter open > mean open in count@.
The primitive $>$ has type $\NN \to \NN \to \NN$ and can be boxed to either @Element@ or @Pure@, but not both.
This means that $\mit{open} > \mit{mean open}$ is a type error.
